Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-08-10 demystify.tex
updates

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[margin=1.5in]{geometry}
\usepackage{soul}
% \usepackage[small,compact]{titlesec} %very powerful
\usepackage[most]{tcolorbox}
% \setsecnumdepth{subsection}
% \setcounter{tocdepth}{3}
\usepackage{enumitem}
\usepackage{epigraph}
\usepackage{cite}
\usepackage{caption}
\captionsetup{font=small}
\usepackage{graphicx}
\usepackage{pdfpages}
\usepackage{hyperref}
\usepackage{wrapfig}
\setlength\intextsep{0pt} % remove extra space above and below in-line float
\usepackage{hyperref}
\hypersetup{
  colorlinks,
  citecolor=black,
  filecolor=black,
  linkcolor=blue,
  urlcolor=blue,
}
\usepackage{booktabs}

@@

\begin{mybox}
This document will be updated regularly to reflect the latest information and updates in the admission process. Its latest version is available at

@@

Most PhD programs will require at least \textbf{two LORs}. A good LoR letter should give details of the applicant's research background and achievements, all of which paint a compelling picture of the candidate to the adcom.
Having a strong letter from an internationally recognized researcher will \emph{greatly strengthen} your application. A strong letter from a well-known researcher can outweigh the lack of publications (or low GPA or unknown school).

@@

\paragraph{Why Self-written Letters are Bad?} Many international students write letters themselves, typically due to the requests of their profs., and have them signed by their profs. Such letters have \textbf{very little value} and are considered weak by reviewers---why can you not even find someone who cares or knows enough about you to write a candid personal reference letter?  Instead of the reference writer talking about you, in this case it is you who write about yourself (and they just sign the letter). Experienced reviewers can easily spot (or suspect) such letters, which can be a \red{red flag}.

\paragraph{Why Generic Letters are Bad?} When the writers do not know much about the applicants (e.g., just taking some course with them or not making any impression to write about), they might write a generic and short letter, which is not useful and also considered weak. This does not mean the professor is not good or does not care about you, but they just do not know you well enough to write a strong letter.
So it might be a good idea to directly ask if the prof. is willing to write a \emph{strong} letter for you. If not, then you should ask someone else.  For example, if a student I don't know well asks me to write a letter for them, I will explicitly tell them I don't know them that well to write much about them, and such a short, generic, and weak letter will not help their case.

@@

Sometimes students would go to great lengths just to get letters from well-known professors in their school (e.g., department head). But as mentioned, if these professors do not know you, their letters would likely be generic and carry little value (sometimes \red{red flags}). Moreover, a top professor at your university might not be well-known to US faculty (see more details in \autoref{sec:your-school}). So save the trouble and get letters from \emph{any} professors/supervisors who know you well and can write a good letter about \emph{your} research ability. It's better to have a good personalized letter about your own research ability from someone who is less well-known than a generic/weak letter from a well-known person.

@@

\paragraph{Why You Want to Help your LOR Writer} As mentioned above, do not write your own letter and generic letters do not give much value. One thing you can do to help your writer to write a strong, customized LoR for you is providing details or unique things about yourself. For example, let them know about your GPA, research and work experiences, papers (if any), or anything you want them to mention. You can also provide them with a draft of your SOP so that they can see what you are saying about yourself and complement that with their own perspective.
Sometimes your writer will explicitly ask you for such information, but if not, you should provide it anyway (especially if you have not interacted with them much or have not done much research with them).

\paragraph{Why You Need to Remind Your Writers} After you submit your application, you should tell your writers about that and let them know they will soon receive an email from the university to submit their letters.  You should also remind them when you submit your application and ask them to submit their letters on time if they haven't done so.  Note that most places only have deadlines for the applicant, but are very flexible with the letter writers (in many cases do not even give them any deadline).  Also, many places do not begin the admission review process right after the deadline and work on application reviews in the next semester (mid-January).

Thus you do want to send reminders because professors can be quite busy and might forget to submit their letters, especially when there is no explicit deadline. However, do not send too many reminders as that can be annoying to the writers.

@@

\begin{center}
  \includegraphics[scale=0.5]{files/c2.png}
\end{center}

@@

In your SOP, focus on research motivation and convince us that you can achieve it through your experience, e.g., published papers. You should also back up your claims with concrete evidence. For example, if you say you have experience with teaching, then show what you did, e.g., undergrad TA or mentoring someone.  If you say you work on a project, then show some results, e.g., paper submitted (or even rejected), achieved certain performance improvement over the state of the art. Also, do not use more than two pages (references can be on the third page). Many SOP samples are \href{https://cs-sop.org/}{available here}.

@@

\epigraph{\vspace{-0.2in} The answer to life's problems aren't at the bottom of a bottle. Heh heh! They're on TV!}{\textsc{The Simpsons}}

@@

\subsection{Selecting and Ranking Schools}\label{sec:selecting-ranking-schools}
\begin{center}
  \includegraphics[scale=0.6]{files/c1.png}
\end{center}

International students often put schools into two bins: (i) top schools that they dream about, and (ii) everything else.  They commonly use rankings from US News, which is not transparent and questionable\footnote{\url{https://cra.org/cra-statement-us-news-world-report-rankings-computer-science-universities/}}.  Sometimes they evaluate based on the reputation of the school's undergrad program or the reputation of the school's non-CS programs such as medical, math, or physics.
They even rank universities based on popular places they know in the US, e.g., California, Texas, and New York.

Instead of these superficial criteria, you should consider the CS program and the research interests of faculty members.
You can learn about these using resources such as \href{https://csrankings.org}{CSRankings.org}. You will be very surprised to learn that a school that you didn't know much about can have very strong research in your topic (and vice versa, a school you thought highly about might have no faculty working in the research field you're interested in). This is also a good way to learn about individual faculty, e.g., who works on what, and well-known CS conferences\footnote{In CS (and probably only in CS), conferences, not journals, are often the main venue to publish research findings (see \href{https://homes.cs.washington.edu/~mernst/advice/conferences-vs-journals.html}{why here}).}. %\autoref{sec:ranking} gives the top 50 CS programs in the US according to CSRankings.

@@

You \emph{do not} need to do a PhD in CS to work in your area of interest. For example, in addition to a traditional CS department, Mason has IST and Cybersecurity departments, both of which have faculty with PhD in CS and work on CS topics (e.g., AI, Security, Robotics).  So it is possible that you still get to do CS research and publish in CS-focused venues even if you're not in a traditional CS program.  It is common to see faculty with PhD in CS in a non-CS department as well as faculty with non-CS PhD in a CS department.

@@

\subsubsection*{Additional Resources}
\begin{itemize}
  \item \href{https://www.cs.columbia.edu/wp-content/uploads/2019/03/Get-Advisor.pdf}{The Definitive ``what do I ask/look for'' in a PhD Advisor Guide}
\end{itemize}
\subsection{Who can serve as a PhD adviser?}\label{sec:faculty-types}

@@

Now that you know a bit about tenured and tenure-track faculty (\autoref{sec:faculty-types}).  Which one should you choose as your adviser? The short answer is that either can be a good adviser, but they have different strengths and weaknesses.

Tenure-track faculty, e.g., assistant professors, are likely young and active in research (they have to, to get tenure). Thus, they will devote more time to work with you and push you to do research and publish. However, they may not have as much experience in managing students and may not have as much funding (yet).

@@

\subsection{Things to avoid when cold emailing a professor}

\begin{itemize}

    \item \textbf{Not customizing the email}. These emails typically talk about the student and not why they are interested in the professor's work. See writing tips in \autoref{sec:contact}
    \item \textbf{Sending the same email to multiple professors}. Sending emails to multiple profs. using BCC or even CC (where we can see all addresses the email is being sent to) will result in no reply or a very harsh one on how unprofessional this is.
    \item \textbf{Too much information}. Keep your email short and to the point. If we are interested, we will ask for more information. 2--3 paragraphs are enough.  Personally I also do not want to see course transcript or CV in the first email.
    \item \textbf{Not being professional}, e.g., do not call the professor by their first name in the first email.  Use Prof. Lastname or Dr. Lastname (\autoref{sec:address}).
    \item \textbf{Not reading the professor's website} and asking questions that are already answered there.
\end{itemize}

@@

\paragraph{Not a good fit and Bad luck}  You could have an excellent profile, but if you are interested in a research area that the program does not have, you might not be admitted.
Similarly, if no faculty who is willing to advise you (e.g., they are on sabbatical or personal leave, do not have sufficient funding, or already have too many students), you will not be admitted.  This is good for you, as you don't want to be in a program where no one can advise you.

@@

Students understandably get excited about their upcoming PhD journey and want to buy new computer equipment and electronics to prepare. However, you should first check with your professor.  They might have funding to buy you computers and other equipment (e.g., software, books, keyboards, headphones, tablets, etc). Many CS programs also provide budget or computer equipment to incoming PhD students (e.g., a laptop).

@@

\chapter{Visa for International Students}\label{sec:visa}

\epigraph{\vspace{-0.2in} I didn’t do it. Nobody saw me do it. You can’t prove anything!}{\textsc{The Simpsons}}

% \begin{center}
%     \includegraphics[scale=0.8]{files/visa.pdf}
% \end{center}

For international students pursuing a PhD in the US, \textbf{F-1} is the main visa that allows a student to study full-time at an accredited institution.  Here are some key points about the F-1 visa:

\begin{itemize}
\item \textbf{Employment:} You are allowed to work on-campus for up to \textbf{20 hours per week during the academic year} (because you still need to take classes) and \textbf{full-time} (typically 40 hrs) during official school breaks (e.g., summer and winter breaks). Off-campus employment requires authorization, which can be obtained through programs like Curricular Practical Training and Optional Practical Training described below.

\item \textbf{Curricular Practical Training (CPT):} CPT allows you to participate in internships or practical training that is an integral part of their academic curriculum. CPT must be related to your field of study and can be full-time or part-time. It requires prior authorization from your university and must be completed before graduation.

\item \textbf{Optional Practical Training (OPT):} OPT provides up to 12 months of work authorization for students before or after completing their degree. For students in STEM fields, there is an additional 24-month extension available. OPT requires prior application and approval from USCIS.

\item \textbf{Full-time Enrollment:} You must maintain full-time enrollment status during the academic year. This means taking a minimum number of credits each semester, as defined by your program. Dropping below full-time status can result in loss of visa status.
\end{itemize}

\includepdf[pages=-]{files/visa.pdf}

\section{For Spouses and Children}

The spouses and children of F-1 visa holders can enter the US under the F-2 visa status. The F-2 visa allows family members to live in the US with the following conditions:

For Spouse: 
\begin{itemize}
\item \textbf{Work Restrictions:} F-2 spouses are not permitted to work in the US.
\item \textbf{Education:} They can study part-time but cannot enroll in full-time degree programs.
\end{itemize}
For Children:
\begin{itemize}
\item \textbf{Education:} F-2 children can attend K-12 schools but cannot pursue higher education full-time.
\item \textbf{Work Restrictions:} Like F-2 spouses, children are not allowed to work under any circumstances.
\end{itemize}
F-2 visa holders must leave the US if the primary F-1 student visa holder loses status or completes their program.

@@

\epigraph{\vspace{-0.2in} I’m not a bad guy! I work hard, and I love my kids. So why should I spend half my Sunday hearing about how I’m going to Hell?}{\textsc{The Simpsons}}

Most of what is written in this document applies to both domestic\footnote{For simplicity domestic means you did your undergrad (or MS) at a US university.} and international students.  However, there are some differences and benefits that domestic students should be aware of and can leverage to improve their chances of admission.

@@

\paragraph{Research Experiences \autoref{sec:research-opportunities}} You might have many opportunities to do research as an undergraduate, e.g., through REUs and internships at your undergrad university.  Highlight such experiences in your application.

@@

\chapter{Research Opportunities}\label{sec:research-opportunities}

\epigraph{\vspace{-0.2in} Kids, you tried your best and you failed miserably. The lesson is: never try.}{\textsc{The Simpsons}}

\begin{center}
    \includegraphics[scale=0.5]{files/phd100404s.png}
  \end{center}

@@

\epigraph{\vspace{-0.2in} Ay caramba!}{\textsc{The Simpsons}}

@@

\epigraph{\vspace{-0.2in} The whole damn system is wrong!}{\textsc{The Simpsons}}

@@

Also thanks to NSF for encouraging faculty to be creative in research and education, which allows me to write this document.

>>>>

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[margin=1.5in]{geometry}
\usepackage{soul}
% \usepackage[small,compact]{titlesec} %very powerful
\usepackage[most]{tcolorbox}
% \setsecnumdepth{subsection}
% \setcounter{tocdepth}{3}
\usepackage{enumitem}
\usepackage{epigraph}
\usepackage{cite}
\usepackage{caption}
\captionsetup{font=small}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{wrapfig}
\setlength\intextsep{0pt} % remove extra space above and below in-line float
\usepackage{hyperref}
\hypersetup{
  colorlinks,
  citecolor=black,
  filecolor=black,
  linkcolor=blue,
  urlcolor=blue,
}
\usepackage{booktabs}

@@

\begin{mybox}
The latest version of this document is available at

@@

Most PhD programs will require at least \textbf{two LORs}. A good LoR letter should give details of the applicant's research background and achievements, all of which paint a compelling picture of the candidate to the adcom.

Having a strong letter from an internationally recognized researcher will \emph{greatly strengthen} your application. A strong letter from a well-known researcher can outweigh the lack of publications (or low GPA or unknown school).

@@

\paragraph{Self-written Letters} Many international students write letters themselves, typically due to the requests of their profs., and have them signed by their profs. Such letters have little value and are considered weak by reviewers---why can you not even find someone who cares or knows enough about you to write a candid personal reference letter?  Instead of the reference writer talking about you, in this case it is you who write about yourself (and they just sign the letter). Moreover, experienced reviewers can easily spot (or suspect) such letters, which can be a \red{red flag}.

\paragraph{Helping your LOR writer} As mentioned above, do not write your own letter. Instead, you can help your LOR writer by providing details or unique things about yourself. For example, let them know about your GPA, research and work experiences, papers (if any), or anything you want them to mention. You can also provide them with a draft of your SOP so that they can see what you are saying about yourself and complement that with their own perspective.
Sometimes your writer will explicit ask you for such information, but if not, you should provide it anyway (especially if you have not interacted with them much or have not done much research with them).

\paragraph{Generic Letters} When the writers do not know much about the applicants (e.g., just taking some course with them or not making any impression to write about), they might write a generic and short letter, which is not useful and also considered weak.  It might be a good idea to directly ask if the prof. is willing to write a \emph{strong} letter for you. If not, then you should ask someone else.  For example, if a student I don't know well asks me to write a letter for them, I will explicitly tell them I don't know them that well to write much about them, and such a short, generic, and weak letter will not help their case.

@@

Sometimes students would go to great length just to get letters from well-known professors in their school (e.g., department head). But as mentioned, if these professors do not know you, their letters would likely be generic and carry little value (sometimes \red{red flags}). Moreover, a top professor at your university might not be well-known to US faculty (see more details in \autoref{sec:your-school}). So save the trouble and get letters from \emph{any} professors/supervisors who know you well and can write a good letter about \emph{your} research ability. It's better to have a good personalized letter about your own research ability from someone who is less well-known than a generic/weak letter from a well-known person.

@@

\paragraph{Reminding Writers} After you submit your application, you should tell your writers about that and let them know they will soon receive an email from the university to submit their letters.  You should also remind them about the deadline and ask them to submit their letters on time.  But do not send too many reminders as that can be annoying to the writers.

Note that most places only have deadlines for the applicant, but are very flexible with the letter writers (in many cases do not even give them any deadline).  Also, many places do not begin the admission review process right after the deadline and work on application reviews in the next semester (mid-January).

@@

\begin{center}
  \includegraphics[scale=0.4]{files/c2.png}
\end{center}

@@

In your SOP, focus on research motivation and convince us that you can achieve it through your experience, e.g., published papers. You should also back up your claims with concrete evidence. For example, if you say you have experience with teaching, then show what you did, e.g., undergrad TA or mentoring someone.  If you say you work on a project, then show some results, e.g., paper submitted (or even rejected), achieved X\% performance improvement over ... Also, do not use more than two pages (references can be on the third page). Many SOP samples are \href{https://cs-sop.org/}{available here}.

@@

@@

\subsection{Selecting and Ranking Schools}\label{sec:selecting-ranking-schools}
\begin{center}
  \includegraphics[scale=0.5]{files/c1.png}
\end{center}

International students often put schools into two bins: (i) very top schools that they dream about, and (ii) everything else.  They commonly use rankings from US News, which is not transparent and questionable\footnote{\url{https://cra.org/cra-statement-us-news-world-report-rankings-computer-science-universities/}}.  Sometimes they evaluate based on the reputation of the school's undergrad program or the reputation of the school's non-CS programs such as medical, math, or physics.
They even rank universities based on popular states they know in the US, e.g., California, Texas, and New York.

You can learn about CS programs and the research interests of faculty using resources such as \href{https://csrankings.org}{CSRankings.org}. You will be very surprised to learn that a school that you didn't know much about can have very strong research in your topic (and vice versa, a school you thought highly about might have no faculty working in the research field you're interested in). This is also a good way to learn about individual faculty, e.g., who works on what, and well-known CS conferences\footnote{In CS (and probably only in CS), conferences, not journals, are often the main venue to publish research findings (see \href{https://homes.cs.washington.edu/~mernst/advice/conferences-vs-journals.html}{why here}).}. %\autoref{sec:ranking} gives the top 50 CS programs in the US according to CSRankings.

@@

@@

\subsubsection*{Additional Resources}
\begin{itemize}
  \item \href{https://www.cs.columbia.edu/wp-content/uploads/2019/03/Get-Advisor.pdf}{The Definitive ``what do I ask/look for'' in a PhD Advisor Guide}
\end{itemize}
\subsection{Types of Faculty: Who can serve as a PhD adviser?}\label{sec:faculty-types}

@@

Tenure-track faculty such as assistant professors are likely young and very active in research (they have to, to get tenure). Thus, they will likely have more time to work with you and push you to do research and publish. However, they may not have as much experience in managing students and may not have as much funding (yet).

@@

You \emph{do not} need to do a PhD in CS to work in your area of interest. For example, in addition to a traditional CS department, Mason has IST and Cybersecurity departments, both of which have faculty with PhD in CS and work on CS topics (e.g., AI, Security, Robotics).  So it is possible that you still get to do CS research and publish in CS-focused venues even if you're not in a traditional CS program.  It is common to see faculty with CS PhD in a non-CS department as well as faculty with non-CS PhD in a CS department.

@@

\paragraph{Not a Good fit and Bad luck}  You could have an excellent profile, but if you are interested in a research area that the program does not have, you will not be admitted.
Similarly, if no faculty is willing to advise you (e.g., they are on sabbatical or personal leave, do not have sufficient funding, or already have too many students), you will not be admitted.  This is actually good for you, as you don't want to be in a program where no one can advise you.

@@

Students understandably get excited about their upcoming PhD journey and want to buy new computer equipment and electronics to prepare. However, you should check with your prof. first.  They might have funding to buy you computers and other equipment (e.g., software, books, keyboards, headphones, tablets, etc). Many CS programs also provide budget or computer equipment to incoming PhD students (e.g., a laptop).

@@

% \section{Additional Resources}
% \begin{itemize}
%   \item \bibentry{nlpphd}
% \end{itemize}

% \chapter{Additional Resources}

% \chapter{Undergraduate Research Opportunities}

% Prospective PhD students are often unaware of the importance of research experience in their application.  This provides both concrete research experiences and potential LOR writers that can talk about your research ability.
% % This is especially true for students from small colleges where research opportunities are limited.
% This section aims to provide some guidance on how to gain research experience as an undergraduate student. %This section is written together Didier, a PhD student at Mason who did his undergraduate study in Rwanda.

% \paragraph{Locally} There are many ways to find research opportunities right in your own institution.  If you did well or like a class, you can check with the professor of that class for research opportunities.  Sometimes professors would mention that they look for undergraduates on their websites, so do check those out! Even if they say they are looking for graduate students, you still should contact them and ask.  You can also ask your academic adviser or other faculty members for suggestions.  Finally, you can also ask your peers who are already doing research.  You can use the method described in \autoref{sec:contact} to contact professors

% In short, the most important thing is that you need to \emph{ask}!  Don't be shy, don't be afraid of rejection, and don't be afraid of being ignored or that you're not good enough.  Just ask!  The worst thing that can happen is that you get a ``no'' or no response.  But you will never know if you don't ask.  And if you get a ``no'', you can always ask again later.  There are students who got rejected the first time but got accepted the second time.  So don't give up!

% \paragraph{Online Research Communities and Open Source Contributions}:
% Places such as Github offer many opportunities for research, even though they might not appear so (e.g., they do not explicitly mention research). Students can explore repositories related to their areas of interest, collaborate with other contributors, and actively engage in open source research projects. By contributing code, fixing bugs, implementing new features, or providing documentation, students can gain practical research experience and interact with experienced developers and researchers in the field.

% \paragraph{Paid or Unpaid}  travel opportunities, publishing opportunities, etc

% \section{Virtual Research Opportunities Beyond Physical Boundaries (written by Didier)}
% \didi{In my experience, prospective PhD students (from Rwanda) are often unaware of the information provided in this document throughout their undergraduate. For this reason they often find themselves unprepared for PhD admission process by the time they finish undergraduate and have to spend a year or more to improve their application profiles. I think it would be beneficial to expand the document to also target students who are not yet ready for the application process to help them improve on all criteria considered during admission. I drafted the short essay on how to gain research experience remotely in order to improve their research profile and potentially get LOR from experts (hopefully faculty)}\\
% In the realm of computer science, research experience plays a crucial role in securing admissions to top-tier PhD programs. However, students from underrepresented small colleges, both in the United States and internationally, often encounter limited research opportunities within their institutions. Thankfully, there exists an alternative pathway for these aspiring researchers to gain valuable research experience, even without being physically present at a university.

% \textbf{Virtual Research Programs}:
% Several universities, organizations, and research institutions offer virtual internships and research programs aimed at providing hands-on research experience. These programs often involve working remotely under the guidance of experienced mentors and collaborating with a team of fellow researchers. For instance, \href{https://docs.google.com/forms/d/1btIwt4HwjyKMOUk-EMy3rbkfWzFxv2lNrMm_zkd0pA4/viewform?edit_requested=true}{UIUC+ Summer Undergraduate Research in Software Engineering}  offers an unpaid remote internship for software engineering students all over the world to collaborate with mentors from University of Illinois at Urbana Champaign.
% Virtual internships offer opportunities to contribute to ongoing research projects, conduct experiments, analyze data, and write research reports. Students can search for such programs through university websites, research institutes, or as of recently chatGPT (or Bard).

% \textbf{Online Conferences and Workshops}:
% Attending online conferences and workshops is another way to gain exposure to cutting-edge research and establish connections with experts in the field. Many conferences now provide virtual participation options, enabling students to access research talks, poster sessions, and panel discussions and sometimes access designated chat rooms or networking events where participants can engage with researchers, ask questions, and seek potential research collaborations. It is beneficial to also create profiles academic collaboration platforms such as ResearchGate. By creating profiles, joining relevant research groups, and participating in discussions, students can connect with established researchers, contribute to ongoing projects, and potentially collaborate on publications or research proposals as they provide access

% In conclusion, for students lacking research opportunities at their small colleges, the virtual realm offers a wealth of possibilities to gain valuable research experience. Engaging with open source projects, joining online research communities, utilizing academic collaboration platforms, participating in virtual internships, and attending online conferences are all avenues to contribute to research, connect with experts, and enhance one's research profile.

@@

Most of what is written in this document applies to both domestic\footnote{For simplicity domestic means you did your undergrad (or MS) at a US university.} and international students.  However, there are some differences and benefits that domestic students should be aware of.

@@

\paragraph{Research Experiences \autoref{sec:research-opportunities}} You might have many opportunities to do research as an undergraduates, e.g., through REUs and internships at your undergrad university.  Highlight such experiences in your application.

@@

\chapter{Research Opportunities}\label{sec:research-opportunities}
\begin{center}
    \includegraphics[scale=0.5]{files/phd100404s.png}
  \end{center}

@@

@@

@@
